# Code-a-thon Project Template

This project is intended to help you apply your skills in R in a way that would be conducive for collaborating or creating reproducible analyses.


## Data Used

Identify a dataset to analyze. Remember that if you share the data you must **not violate any privacy restrictions**

If you don't have your own data, options for places to find data are listed on https://daseh.org.

You might start by setting up an R Project, importing your data, previewing the data, etc. 

This dataset was collect from NCBI.

```{r}
install.packages("tidyverse")
install.packages("ggplot2")
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("edgeR")

library(tidyverse)
library(ggplot2)
library(edgeR)
library(RUVSeq)
library(RColorBrewer)
library(openxlsx)
```

**describe** what your data looks like and **where** you got it. Identify what the variables and samples are, or point to a data dictionary. Describe how the data was originally created. 


# Data Cleaning/Wrangling

Perform any data subsetting, cleaning, or manipulation methods that were described in this course on your data. Examples are: renaming the columns, recoding values, reshaping the data, filtering the data etc. 

You might also research other methods to include here, especially if your data is very unique.

```{r}
transporter <- read.delim("C:/Users/SaraS/OneDrive/Documents/AmeliaS Code a thon - 20240731/placenta.txt", row.names="Symbol")
view(transporter)

transporter_filter <- dplyr::filter(transporter, T1_1 >8 & T1_2 >8 & T1_3 >8 & T1_4 >8 & T3_1 >8 & T3_2 >8 & T3_3 >8 & T3_4 >8)

# T1 <- dplyr::select(transporter_filter, T1_1, T1_2, T1_3, T1_4)
# T3 <- dplyr::select(transporter_filter, T3_1, T3_2, T3_3, T3_4)



```


# Data Visualization

Create some visualizations of your data using the esquisse app or the ggplot2 package.

```{r}
# Separate the data into T1 and T3
# test for differential expression in each gene using QL F-test
sample_name <- c("T1_1", "T1_2", "T1_3", "T1_4", "T3_1", "T3_2", "T3_3", "T3_4")

group <- c("first", "first", "first", "first", "third", "third", "third", "third")

group.df <- data.frame(sample_name, group)

# design matrix
design <- model.matrix(~0+group) # either 0 or 1
colnames(design) <- levels(group) 
colnames(design) <- c("first", "third")
design
```

# dispersion estimation
```{r}
y <- estimateDisp(y, design, robust = TRUE)
plotBCV(y)
```
```{r}
fit <- glmQLFit(y, design, robuts = TRUE)
head(fit$coefficients)
```
```{r}
plotQLDisp(fit) # better corrected dispersion
summary(fit$df.prior)
```


```{r}
library(edgeR)

y <- DGEList(transporter_filter, group = group, 
        genes = row.names(transporter_filter))
head(y)

# pch = c(0, 1) # only two groups

color <- c("darkgreen", "darkgreen", "darkgreen", "darkgreen", "yellow", "yellow", "yellow", "yellow")
plotMDS(y, pch = 19, col = color[group]) # each group get a different symbol
# after we add the color, the datapoint is not showing up in the plot - need to figure this out 

plotMD(y, column = 1)

# top PC is the one that maximize the differences !!!
```


```{r}
# differential analysis 

con <- makeContrasts(transporter_filter = first - third, levels = design)
con 
# to see the dimentions of the dataset
dim(design)
dim(transporter_filter)
```


```{r}
res <- glmQLFTest(fit, contrast = con)
topTags(res)
```


```{r}
# to control for the FDR 
is.de <- decideTestsDGE(res)
summary(is.de)

# plot it 

plotMD(res, status = is.de, values = c(1, -1), col=c("red", "blue"), 
       legend = "topright")
```


```{r}
# to control the fold change above 1.5 to be significant change
threshold <- glmTreat(fit, contrast = con, lfc = log2(1.5))
topTags(threshold)

# control for FDR
is.de <- decideTestsDGE(threshold)
summary(is.de)

# plot it 

plotMD(threshold, status = is.de, values = c(1, -1), col=c("red", "blue"), 
       legend = "topright")
```

```{r}
# maeka volcano plot
library(RColorBrewer)
install.packages("ggrepel")
library(ggrepel)
view(threshold)
threshold_table <- threshold$table
ggplot(data = threshold_table, aes(x = logFC, y = -log10(PValue)))+
  geom_point()
```


```{r}
# Add threshold lines

library(ggplot2)

#ggplot(threshold_table) +
 #aes(x = logFC, y = PValue) +
 #geom_point(colour = "#112446") +
 #theme_minimal()


# ggplot(data = threshold_table, aes(x = logFC, y = -log10(PValue))) +
#  geom_vline(xintercept = c(-1.5, 1.5), col = "darkgreen", linetype = 'dashed') +
 # geom_hline(yintercept = -log10(0.05), col = "darkgreen", linetype = 'dashed') + 
 # geom_point() 



threshold_table$diffexpressed <- "NO"

threshold_table$diffexpressed[threshold_table$logFC > 1.5 & threshold_table$PValue < 0.05] <- "UP"
threshold_table$diffexpressed[threshold_table$logFC < -1.5 & threshold_table$PValue < 0.05] <- "DOWN"
p <- ggplot(data=threshold_table, aes(x=logFC, y=-log10(PValue), col=diffexpressed)) + geom_point() + theme_minimal()

p2 <- p + geom_vline(xintercept=c(-1.5, 1.5), col="red") +
        geom_hline(yintercept=-log10(0.05), col="red")

mycolors <- c("blue", "red", "black")
names(mycolors) <- c("DOWN", "UP", "NO")
p3 <- p2 + scale_colour_manual(values = mycolors)
p3
```


```{r}
#label the points 
threshold_table2 <- threshold$genes
threshold_table$delabel <- NA
threshold_table$genes <- threshold_table2
threshold_table$delabel[threshold_table$diffexpressed != "NO"] <- threshold_table$genes$genes[threshold_table$diffexpressed != "NO"]


ggplot(data=threshold_table, aes(x=logFC, y=-log10(PValue), col=diffexpressed, label=delabel)) + 
    geom_point() + 
    theme_minimal() +
    geom_text()

library(ggrepel)
ggplot(data=threshold_table, aes(x=logFC, y=-log10(PValue), col=diffexpressed, label=delabel)) +
        geom_point() + 
        theme_minimal() +
        geom_text_repel() +
        scale_color_manual(values=c("blue", "black", "red")) +
        geom_vline(xintercept=c(-0.6, 0.6), col="red") +
        geom_hline(yintercept=-log10(0.05), col="red")


```


```{r}
# heatmap clustering
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("ComplexHeatmap")

library(ComplexHeatmap)

logCPM <- cpm(y, prior.count = 2, log = TRUE)
rownames(logCPM) <- y$genes$genes
colnames(logCPM) <- paste(y$samples$group, 1:4, sep = "-")

o <- order(threshold$table$PValue)
logCPM <- logCPM[o[1:30],]

#  we scale each row (each gene) to have mean zero and standard deviation one:
logCPM <- t(scale(t(logCPM)))

library(gplots)
col.pan <- colorpanel(100, "blue", "white", "red")
heatmap.2(logCPM, col = col.pan)


Heatmap(logCPM, col = col.pan, 
        name = "HeatMap of the top 30 most DE genes between the First and Third Trimester",
        row_title = "genes", 
        column_title = "gestational stages", 
        column_title_side = "bottom", 
        heatmap_width = unit(100, "mm"), 
        heatmap_height = unit(200, "mm"))

# Heatmap function in ComplexHeatmap
# Heatmap(matrix, col, name, na_col = "grey", color_space = "LAB", rect_gp = gpar(col = NA), border = NA, border_gp = gpar(col = "black"), cell_fun = NULL, layer_fun = NULL, jitter = FALSE, row_title = character(0), row_title_side = c("left", "right"), row_title_gp = gpar(fontsize = 13.2), row_title_rot = switch(row_title_side[1], left = 90, right = 270), column_title = character(0), column_title_side = c("top", "bottom"), column_title_gp = gpar(fontsize = 13.2), column_title_rot = 0, cluster_rows = TRUE, cluster_row_slices = TRUE, clustering_distance_rows = "euclidean", clustering_method_rows = "complete", row_dend_side = c("left", "right"), row_dend_width = unit(10, "mm"), show_row_dend = TRUE, row_dend_reorder = is.logical(cluster_rows) || is.function(cluster_rows), row_dend_gp = gpar(), cluster_columns = TRUE, cluster_column_slices = TRUE, clustering_distance_columns = "euclidean", clustering_method_columns = "complete", column_dend_side = c("top", "bottom"), column_dend_height = unit(10, "mm"), show_column_dend = TRUE, column_dend_gp = gpar(), column_dend_reorder = is.logical(cluster_columns) || is.function(cluster_columns), row_order = NULL, column_order = NULL, row_labels = rownames(matrix), row_names_side = c("right", "left"), show_row_names = TRUE, row_names_max_width = unit(6, "cm"), row_names_gp = gpar(fontsize = 12), row_names_rot = 0, row_names_centered = FALSE, column_labels = colnames(matrix), column_names_side = c("bottom", "top"), show_column_names = TRUE, column_names_max_height = unit(6, "cm"), column_names_gp = gpar(fontsize = 12), column_names_rot = 90, column_names_centered = FALSE, top_annotation = NULL, bottom_annotation = NULL, left_annotation = NULL, right_annotation = NULL, km = 1, split = NULL, row_km = km, row_km_repeats = 1, row_split = split, column_km = 1, column_km_repeats = 1, column_split = NULL, gap = unit(1, "mm"), row_gap = unit(1, "mm"), column_gap = unit(1, "mm"), show_parent_dend_line = ht_opt$show_parent_dend_line, heatmap_width = unit(1, "npc"), width = NULL, heatmap_height = unit(1, "npc"), height = NULL, show_heatmap_legend = TRUE, heatmap_legend_param = list(title = name), use_raster = NULL, raster_device = c("png", "jpeg", "tiff", "CairoPNG", "CairoJPEG", "CairoTIFF", "agg_png"), raster_quality = 1, raster_device_param = list(), raster_resize_mat = FALSE, raster_by_magick = requireNamespace("magick", quietly = TRUE), raster_magick_filter = NULL, post_fun = NULL))
```


```{r}
# KEGG pathway analysis 
keg <- kegga(threshold, species = "Hs")
topKEGG(keg, n = 15, truncate = 34)

```
# Data Analysis

Perform a **simple analysis** of your data. This can involve summarizing the data to describe aspects about it (quartiles, means, range etc.) or a simple statistical test. 

Don't forget to describe what analysis you performed and why. Provide some simple **interpretation** about what your analysis might indicate about your data.


# Please include additional sections as desired.

```{r}

```


# Versions

Include some information about the version of R and the packages you are using by running sessionInfo().
1


```{r}
sessionInfo()
```

Render your document to an html or pdf using the "Knit" button.
